---
title: Estimating the effect of luxury cabin on the chances to survive the sinking
  of the Titanic
author: "Florian Benkhalifa, Camilla Bischofberger, Max Röcker"
date: "3 5 2020"
output:
  pdf_document:
    keep_tex: true
    fig_caption: yes
header-includes:
   - \usepackage{dcolumn}  
   - \usepackage{float}  
   - \usepackage{flafter}  
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, fig.pos = "h", out.extra = "")
library(datasets)
# Data wrangling
library(tidyverse)
library(lubridate)
library(reshape)
library(stringi)

# Data viz
library(plotly)
library(scales)
library(ggthemes)
library(cowplot)
library(stargazer)
library(skimr)
library(ggmosaic)
library(ggpubr)

# Data presentation
library(knitr)
library(stargazer)

# Data analysis
library(tidymodels)
library(broom)
library(glmnet)
library(rsample)
library(furrr)
```

```{r preparation, include = FALSE}

# Convert array to 2-dimensional format
mydata <- Titanic %>% as_tibble() %>% print(n=50)

# Convert all except n to factors which are easier to handle later on
mydata <- mydata %>% mutate_at(vars(-n), as.factor) 

# Untable the dbl which makes it easier to use with summarize functions
mydata <- untable(df = mydata %>% select(-n), 
                  num = mydata %>% pull(n))

# Prepare
mydata_rec <-
  recipe(Survived ~ ., data = mydata) %>% 
  step_dummy(all_predictors()) %>% 
  step_interact(terms = ~ starts_with("Sex"):starts_with("Age")) %>% 
  step_zv(all_predictors()) %>% 
  prep() %>% 
  juice()


load("./cv_model/cv_results.rda")

```

## Introduction

The purpose of this paper is to assess the effect of the economic status (class) of the cabin on the chances of survival as a passenger on the Titanic. The fate of the R.M.S. Titanic and its passengers has captured the attention of the whole world. There is consensus that the chances of getting the few places in the limited number of lifeboats and surviving the cold waters of the North Atlantic differed among social groups. Remarkably, women rather than men survived the disaster [@Hall1986]. This was apparently due to the long time of the sinking (2.6h) that allowed for social norms such as ‘women and children first’ to be established @Frey2010. However, this did not apply across cabin classes: Eyewitness reports suggest that there was a structural disadvantage and discrimination of third-class passengers [@Diekmann2012]. Accordingly, survivors were rather from the upper cabin classes than from lower classes [@DixonGriffiths2006, @Frey2011, @Hall1986]. Therefore, ex ante, the effect seems to be clear – the lower the class of a passenger, the smaller the chances for survival. This inequality would constitute an appalling injustice and thus, necessitates a factual verification. Fortunately, we received passenger data from the British Board of Trade that enables us to assess this effect. 
We perform a thorough analysis in the empirical framework of a logistic regression. We find that the relationship between class and chances of survival is best described by including additional variables and an interaction term. We conclude that it is likely that the effect on the survival rate might be a combination of several factors, with economic status of the cabin being a significant one. The results of this study should inform our struggle for safe passages for everyone and workers rights worldwide.


## Data and descriptives

The dataset “Titanic” contains information about the survival status of a person, gender, the economic class and the age of each passenger. All the variables are categorical, with only the economic class having four categories - the remaining are binary. Figure \ref{fig:categories}
```{r categories, fig.align="center", echo = FALSE, fig.cap="Descriptive figures"}

# 3 Produce a table which helps to grasp differences within groups more easily
p1 <- mydata %>%
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "attribute",
    names_repair = "universal"
  ) %>% ggplot(aes(x = attribute, group = variable)) +
  geom_bar(aes(y = (..count..)))+
  ylab("n") +
  labs(title = "Histograms per variable") +
  facet_wrap(~variable, scales = "free_x") +
  theme_bw() +
  theme(text = element_text(size=10))

# 4 Compute survival rate which will be included in next plot
survival_rate <- mydata %>% 
  count(Survived, .drop = FALSE) %>% 
  mutate(freq_per_class = n/sum(n))

# 5 Convert data to long format for facet plotting in ggplot
p2 <- mydata %>% 
  pivot_longer(-Survived, 
               names_to = "variable", 
               values_to = "attribute", 
               names_repair = "universal") %>% 
  mutate_if(is.character, as.factor) %>% 
  ggplot(aes(x = attribute, fill = Survived)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent) +
  ylab("Survival Rate") +
  theme(axis.text.x = element_text(angle=45, hjust=1))+
  scale_fill_manual(values = c("gray25", "gray")) +
  geom_hline(yintercept = survival_rate$freq_per_class[2], col = "white", lty = 2, size = 1) +
  facet_wrap(~variable, scales = "free_x") +
  labs(title = "Survival outcome per category") +
  theme_bw()+
  theme(legend.position="bottom", text = element_text(size=10))

## !!!!More descriptive stats can be found in the appendix!!!!

ggarrange(p1, p2, nrow = 1, widths = c(1, 1.3))

```

In total there are $2201$ observations in the dataset. A simple summary statistic denoting the marginal distributions of the variables is illustrated in figure xxx. 
In order to get a preliminary grasp of the relation between the regressors and the target variable, several stacked barplots are provided in Figure 1. They graphically indicate differences in the survival rate between multiple groups. The white dotted line marks the overall mean survivals and facilitates to detect conspicious groups.
The graph underlines three provisional findings: 
- Children had a higher chance to survive than adults, 
- survival rate diminishes with economic class and 
- females were more likely to survive the crash than males. 

From a first glance the preliminary findings seem to be linked to traditional shipping conducts. Findings 1 and 3 mostly follow from the “Women and children first” policy which priorises the lives of women and children in life threatening situations. In turn it is not surprising that male passengers reveal a low survival rate of less than 20%. Finding (2) would lead into the direction of the presumption about inequality among the classes described in the introduction. Thus, validating finding 2 is of crucial interest in the next section.

## Empirical strategy and results

In this section we estimate the causal effect of the economic status of the cabine on the chances of survival. We start with the simplest form of logistic regression model applicable to the data which takes the form

$$P(Survived = 1|x)= G(\beta_o + \beta_1Class1 + \beta_2Class2 + \beta_3Class3) + u$$ 
where $i$ denotes the i-th individual and $\beta_0$, $\beta_1$, $\beta_2$ and $\beta_3$ are the unknown coefficients to be estimated. $u$ is the error of the regression and remains unobserved. The role of $G$ is to keep the probability $P(Survived = 1)$ in between zero and one. The effect of interest is captured in equation (1) by the beta coefficients which denote the increase of log odds when the survival probability of survival of a passenger belonging to a lower class is compared to a higher one. Because the coefficients are problematic to be interpreted directly, the marginal effect,

We abstained from constructing this model via a linear relationship, as assumed by OLS, since for some range of the covariates the estimated conditional probability may fall outside of $[0,1]$. Clearly, such estimations violate the premise of probability functions. Further, a linear probability model implies that a ceteris paribus increase in $x_i$ always corresponds with a constant change of $P[Survival=1|x_i]$ regardless of the initial value of $x_i$. This also implies that we can get $P[Survival=1|x_i]$ greater or less than $[0,1]$ if we increase or decrease the dependent variable continually. Particularly when trying to estimate partial effects for extreme values of $x$, this should cause concerns.

Here, the variable $Survived$ is a binary variable indicating whether a person survived $Survived = 1$ or not. However, not only the dependent variable is categorical, all explenatory are categorical as well. To account for this specificity with regard to this paper's purpose, namely to investigate the effect economic class on the chances of survival, the explanatory variables are converted into three dummy variables. This allows  since the magnitude of switching from Class 1 to Class 2, for instance, does not convey useful information, only ordinal information.This ensures that the beta coefficients are more flexible, since they are not bound to increase in a linear way for every class. We have chosen to not make a dummy variable for “Class 1” as the effect of being part of the first class is depicted in $\beta_0$ while in this case all the other “Class” variables are 0. Including too many dummy to describe a group, can lead to the so-called dummy variable trap @Wooldridge [p.213]. Thus, the base group in this model is the first class. Consequently, $\beta_1$, $\beta_2$ and $\beta_3$ depict the increase or decrease of the survival probabilities in comparison with the base group.
The estimation results of this simple model are displayed in the first column of table xxx

```{r cv, results = 'asis', fig.pos = "H"}
# Create table via stargazer, tehr
cv_results <-  model_metrics %>% select( c(accuracy, kap)) %>% imap(~c(paste0("cv.",.y), round(.x, digits = 3)))

# Create the table
model_metrics %>% 
  pull(fit) %>% 
  exec(stargazer, ., title = "Results", align = TRUE,
       omit.stat=c("f", "ser"), order=c("Constant",
                                        "Class_X2nd",
                                        "Class_X3rd",
                                        "Class_Crew",
                                        "Sex_Male",
                                        "Age_Child",
                                        "Sex_Male_x_Age_Child"),
       add.lines= cv_results,
       font.size = "small",
       column.sep.width = "-24pt",  header=FALSE,  table.placement = "t")
```

The estimation results of this simple model are displayed in the first column of table X. The Constant - b0=0.51 - is the only estimate which is positive. Thus, this analysis suggests that the first class has the highest chance of survival compared to the other classes and the crew.  The estimate for the crew class is the lowest with -1.66 which suggests that being a member of the crew class has the strongest negative effect on the chances to survive. However, the estimate for the third class is very close to the one of the crew with -1.6. So being a passenger in the third class or being a crew member reduces the chances of survival almost equally compared to the first class passengers (base group). This matches what we see in Table 1 - the survival rate of the third class and the crew class are almost the same around 25% and the first class has the highest survival rate. 
As the estimates from our simple regression model are all significant in the 99% confidence bound with p-values virtually equal to zero, we can reject the Null hypotheses that bi=0.
We now perform the model diagnostics. The important questions are whether the estimates are credible and whether the assumptions behind the simple model hold. In order to interpret the variable in a causal manner, this simple randomized model works under two main assumptions: First, that on average the error is zero $(E[u_i = 0])$ and second, that there is randomization $(x \bot u)$ and thus, no correlation between our error term and the Class variables (exogeneity). As we will see further down in our analysis, these assumptions do not hold for the simple model.
The error term might contain the other variables from the data which are gender and age. Moreover, factors contained in the error term could be body weight, body fat, physical condition, swimming abilities, intelligence, clothing, number of family members on the ship, location in the ship, the distribution of flotsam and pure chance. However, we do not have data on these factors. Our data doesn’t come from a randomized experiment but we could think of a randomized experiment in order to extract the effect of class. We would randomly pick 1000 people representative for all strands of society, assign them randomly to different cabin classes on an ocean liner and let the ship sink in the Atlantic. Fortunately, this experiment will most probably never take place.

As we have binary variables, there is no point in adding polynomials of higher degree to our model as this doesn’t help in explaining the effect of interest. So we are starting off by separately adding the dummy variables for sex (column 2) and age (column 3) to the model. It is very interesting to see how the estimates in column 2 change in comparison to column 1. The estimates are still all significant. By adding the variable sex (male=1), the constant which stands for the base group of the first class, increased from 0.5 to 2.1. The estimates for the second and third class haven’t changed that much. The crew class estimate still indicates a negative effect on the survival chance compared to the first class. However, the estimate changed from -1.7 to -0.9. In column 2, the estimate for being male is -2.4. To understand this, let us consider a male first class passenger. His chance of survival would be 2.068-2.421= -0.353. If on the other hand, we consider a female first class passenger, her survival change would be 2.068 which is much higher as the dummy-variable for sex is 0. The lowest survival chance would a male third class passenger have with -2.011. 
Adding the age-dummy to the simple model 1 doesn’t change the estimates as much as adding a sex-dummy as we see in column 3. The estimate for age (child=1) is also significant and indicates a positive effect on the chances of survival. In adding both dummy variables, we observe in column 5 the same changes to the estimates which we observed from only adding the sex variable. The estimates in column 5 are all significant with a very low p-value. The change in the constant and the estimates for the classes, we can assume that there was omitted variable bias in the simple model which means that the above assumption of exogeneity doesn’t hold. 
Further, we add an interaction term to our model. There is a myriad of options for interaction terms but we have opted to display the interaction of sex (male=1) and age (child=1). Columns 4, 6 and 7 show the models where we have added the interaction term separately and in combination with either the sex dummy or the age dummy. Depending on the combination, the estimate for the interaction term differs greatly and is least significant and negative for model 7 where we estimate it together with the age dummy. Column 8 depicts the estimates for the model where we have included all the dummy variables and the interaction term. The estimates from column 8 and column 6 are almost the same but including all the variables decreases the bias and the chance of endogeneity in the model. 

Clearly this dataset does not deliver unbiased effects of the ecomomic classe. In order to extract the effect of class, we could think of a randomized experiment. Therefore, we would randomly pick 1000 people representative for passengers who prefer to travel by a liner. For instance, past reservations on different liners across the world could be pooled into a dataset, and 1000 people are randomly chosen and informed that they won a free a cruise in a lottery. Once they arrive on the liner, which has also been randomly chosen, they are randomly assigned to different cabin classes holding all other factors such as ship policy equal. To eventually test how the economic class affects the survival chance, the ship is directed towards 
```{r, results = 'asis', echo = FALSE}
model_metrics$fit[[1]] %>% anova(test = "Chisq")%>% as.matrix %>% stargazer(., title = "Chi-Squared for Model (8)", font.size = "small", header=FALSE)

```

The likelihood ratio tests indicates that adding each variable to the regression clearly improved the model.

```{r, results = 'asis', echo = FALSE}
model_metrics$fit[[8]] %>% anova(test = "Chisq") %>% as.matrix %>% stargazer(., title = "Chi-Squared for Model (1)", font.size = "small", header=FALSE)

```
The same is true for our complex model which increases over time.

```{r, results = 'asis', echo = FALSE}
anova(model_metrics$fit[[1]], model_metrics$fit[[8]],  test = "Chisq") %>% as.matrix %>% stargazer(., title = "Comparison of model (1) and (2) via a likelihood ratio test", font.size = "small", header=FALSE)

```
We finally make a likelihood ratio test to find out if the complex model really outperforms the simple model. The p-value is significant indicating that the complex model increased the performance.

## Conclusion

In summary, the economic status of the cabin (class) appears to have an effect on the survival chances. The positive effect which a passenger in the first class enjoys on its survival chances decreases with economic class (crew included). Thus, our findings concur with the literature. In consequence, we suggest a policy change that makes the Atlantic passage safe for all passengers, regardless of class. This could include raising the number of lifeboats and life vests to a number that reflects the passenger numbers and ensuring that lifeboats can easily be accessed from different cabin class areas.
